---
title: "ACCESS-S2 eval"
format: 
  revealjs:
    fig-width: 12
    fig-height: 6
execute: 
  cache: refresh
---

```{r setup, cache=FALSE}
library(metR)
library(ggplot2)
library(data.table)
library(patchwork)
theme_set(theme_minimal())


set_day <- function(date) {
  lubridate::day(date) <- 01
  as.Date(date)
}
data <- here::here("data/data_derived")
```

# HINDCAST

```{r read_data}

read_extent <- function(extent, area, name = "aice", range = NULL) {
  if (!is.null(range)) {
    range <- list(time = range)
  }
  extent <- ReadNetCDF(extent, vars = c(extent = name),
                       subset = range)
  
  area <- ReadNetCDF(area, vars = c(area = name),
                     subset = range)
  
  extent[area, on = .NATURAL] |>  
    _[, time := as.Date(time)] |> 
    _[, let(lat = NULL, lon = NULL)] |> 
    _[, tidyr::complete(.SD, time = seq(min(time),
                                        max(time),
                                        "1 month"))] |> 
    as.data.table() |> 
    melt(id.vars = c("time"))
}

access <- read_extent(file.path(data, "access_extent.nc"), 
                      file.path(data, "access_area.nc"))

range_access <- range(access$time)

hadisst <- read_extent(file.path(data, "hadisst_extent.nc"), 
                       file.path(data, "hadisst_area.nc"),
                       name = "sic",
                       range = range_access)


nsidc <- read_extent(file.path(data, "nsidc_extent.nc"), 
                     file.path(data, "nsidc_area.nc"),
                     name = "cdr_seaice_conc_monthly",
                     range = range_access)

```

## Antarctic Sea Ice Extent/Area

```{r}
extents <- rbind(access = access, hadisst = hadisst, nsidc = nsidc, idcol = "model") |>
  _[value == 0, value := NA] |>  # Problems with NSIDC
  _[, value := Anomaly(value, year(time) %between% c(1981, 2011), na.rm = TRUE), 
    by = .(model, variable, month(time))] |> 
  _[, model := forcats::fct_relevel(model, "access", after = Inf)]


cors <- extents |>
  _[, widyr::pairwise_cor(.SD, model, time, value, use = "complete.obs",
                          diag = FALSE, upper = FALSE), 
    by = variable] |> 
  _[, paste0("r(", item1, ", ", item2, ") = ", scales::number(correlation, accuracy = 0.01)) |> 
      paste0(collapse = "\n"), 
    by = variable] 

extents |> 
  copy() |> 
  # _[model == "hadisst", extent := extent/2] |> 
  ggplot(aes(time, value)) +
  geom_line(aes(color = model), linewidth = 0.7) +
  
  geom_smooth(aes(color = model), method = "lm",
              formula = y ~ x +
                I(pmax(x - as.numeric(as.Date("2005-12-01")), 0)) +
                I(pmax(x - as.numeric(as.Date("2012-12-01")), 0)),
              n = 100
  ) +
  
  scale_y_continuous("Extent/area (million km²)",
                     labels = scales::label_number(scale = 1e-12)) +
  scale_x_date(NULL, expand = c(0, 0)) +
  scale_color_manual(NULL, 
                     values = c(access = "black", 
                                hadisst = "#3584e4", 
                                nsidc = "#e66100"),
                     labels = c(access = "ACCESS-S2",
                                hadisst = "HadISST",
                                nsidc = "NSIDC CDRV4")) +
  geom_text(data = cors, aes(x = as.Date("1990-01-01"), y = 4*1e12, label = V1), vjust = 1) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  facet_wrap(~ variable, ncol = 1)

```


## Antarctic Sea Ice Extent/Area

```{r}
extents |> 
  dcast(time + variable ~ model, value.var = "value") |> 
  melt(id.vars= c("time", "variable", "access"), variable.name = "model") |> 
  ggplot(aes(value, access)) +
  geom_point(size = 0.4) +
  geom_smooth() +
  scale_y_continuous("Extent/area (million km²)",
                     labels = scales::label_number(scale = 1e-12)) +
  scale_x_continuous("Extent/area (million km²)",
                     labels = scales::label_number(scale = 1e-12)) +
  facet_grid(variable~model)
```

## Spatial correlation of anomalies

```{r}

correlations <- rbind(nsidc = ReadNetCDF(file.path(data, "correlation_space_nsidc.nc"), 
                                         vars = c(correlation = "cdr_seaice_conc_monthly")) |> 
                        na.omit() ,
                      hadisst = ReadNetCDF(file.path(data, "correlation_space_hadisst.nc"), 
                                           vars = c(correlation = "sic")) |> 
                        na.omit(),
                      idcol = "dataset") 

# correlations |> 
#   ggplot(aes(time, correlation)) +
#   geom_hline(yintercept = 0, linetype = 3) +
#   geom_line(aes(color = dataset), linewidth = 0.7) +
#   scale_color_manual(NULL, 
#                      values = c(access = "black", 
#                                 hadisst = "#3584e4", 
#                                 nsidc = "#e66100"),
#                      labels = c(access = "ACCESS-S2",
#                                 hadisst = "HadISST",
#                                 nsidc = "NSIDC CDRV4")) +
#   theme_minimal() +
#   scale_y_continuous("Correlation of anomalies") +
#   scale_x_datetime(NULL, expand = c(0, 0)) +
#   theme(legend.position = "bottom") 


correlations |> 
  ggplot(aes(month(time), correlation)) +
  geom_hline(yintercept = 0, linetype = 3) +
  geom_violin(aes(group = month(time), color = dataset)) +
  ggforce::geom_sina(aes(group = month(time), color = dataset)) +
  scale_y_continuous("Correlation of (spatial) anomalies") +
  scale_x_continuous(NULL, breaks = 1:12, expand = c(0, 0), labels = month.abb) +
  scale_color_manual(NULL, guide = "none",
                     values = c(access = "black", 
                                hadisst = "#3584e4", 
                                nsidc = "#e66100"),
                     labels = c(access = "ACCESS-S2",
                                hadisst = "HadISST",
                                nsidc = "NSIDC CDRV4")) +
  facet_wrap(~dataset, ncol = 1, 
             labeller = labeller(dataset = c(hadisst = "HadISST",
                                             nsidc = "NSIDC CDRV4")))
```

## Time correlation of anomalies

```{r}
map_antarctica <- map_data("world2") |> 
  as.data.table() |> 
  _[lat < -50] |> 
  geom_polygon(data = _, inherit.aes = FALSE, aes(long, lat, group = group), 
               colour = "black", linewidth = 0.3, fill = NA)

correlation_time |> 
  _[, month := month(time)] |> 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = correlation)) +
  scale_fill_divergent("Correlation") +
  scale_y_latitude(limits = c(-90, NA), labels = NULL) + 
  scale_x_longitude() +
  map_antarctica +
  coord_polar() + 
  facet_wrap(~ month, labeller = labeller(month = setNames(month.abb, 1:12))) 
```


---

Why is the correlation so high in the anomalies, but so low in the area? 

---

It's not really that high

```{r}
correlation_time |> 
  _[correlation != 1] |> 
  _[, month := lubridate::month(time, label = TRUE, abbr = TRUE)] |> 
  ggplot(aes(correlation)) +
  geom_density(aes(weight = cos(lat*pi/180))) +
  facet_wrap(~month)
```

# Forecats

# Forecats 😸

## Initial conditions vs forecast

```{r}
files <- list.files(file.path(data, "correlations"), full.names = TRUE) 

read_correlations <- function(files) {  
  dates <- utils::strcapture("mi_aice_(\\d{8})_e(\\d{2})_(\\w+).nc", basename(files),
                             proto = list(time_forecast = character(),
                                          member = character(),
                                          type = character())) |> 
    as.data.table() |> 
    _[, time_forecast := as.Date(time_forecast, format = "%Y%m%d")] |> 
    _[]
  
  read <- function(i) {
    ReadNetCDF(files[[i]], vars = c(value = "aice")) |> 
      _[, let(lat = NULL, lon = NULL, time = set_day(time))] |> 
      _[] |> 
      _[, let(time_forecast = dates$time_forecast[[i]],
              member = dates$member[[i]],
              type = dates$type[[i]])] |> 
      _[] 
  }
  
  lapply(seq_along(files), read) |> 
    rbindlist() 
}

geom_cross <- function() {
  list(geom_hline(yintercept = 0), 
       geom_vline(xintercept = 0))
}
```

```{r}
cors <- read_correlations(list.files(file.path(data, "correlations"), full.names = TRUE)) |> 
  _[, measure := "correlation"] |> 
  tidyfast::dt_pivot_wider(names_from = type, values_from = value) |> 
  _[, let(skill = (forecast - persistence)/(1 - persistence))] 


rmse <- read_correlations(list.files(file.path(data, "rmse"), full.names = TRUE)) |> 
  _[, measure := "rmse"] |> 
  _[value < 0.01, value := NA] |>    # A few forecast have very low values
  tidyfast::dt_pivot_wider(names_from = type, values_from = value) |> 
  _[, let(skill = (forecast - persistence)/(0 - persistence))] 

```

```{r}
skill <- rbind(cors, rmse) |> 
  _[, let(persistence = NULL)]

initial <- skill[time == time_forecast, .(time_forecast, member, measure, initial = forecast)]

skill <- skill[time != time_forecast] |> 
  _[initial, on = c("time_forecast", "member", "measure")] |> 
  _[, let(lead = months_difference(time_forecast, time),
          verification_season = season(time),
          forecast_season = season(time_forecast))] 
```


```{r}
rbind(cors, rmse) |> 
  _[month(time_forecast) %in% c(02, 05, 08, 11)] |> 
  melt(id.vars = c("time", "time_forecast", "member", "measure"),
       measure.vars = c("forecast", "persistence")) |> 
  _[, lead := months_difference(time_forecast, time)] |> 
  _[measure == "rmse"] |> 
  _[lead <= 6] |> 
  _[, .(value = mean(value, na.rm = TRUE)), 
    by = .(lead, month(time_forecast), measure, variable)] |> 
  ggplot(aes(lead, value)) +
  geom_line(aes(color = variable)) +
  facet_grid(month ~ measure)
```


```{r}
skill |> 
  _[measure == "rmse"] |> 
  ggplot(aes(skill)) +
  geom_density(aes(color = factor(lead))) +
  facet_wrap(lubridate::month(time, label = TRUE)~.)
```




```{r}
skill |> 
  _[measure == "rmse"] |> 
  na.omit() |> 
  ggplot(aes(initial, forecast)) +
  
  geom_point(size = 0.4) +
  geom_smooth(method = "lm", color = "#1c71d8", se = FALSE) +
  # geom_line(data = lines, linewidth = 0.9, color = "#e66100") +
  scale_x_continuous("Initial conditions") +
  scale_y_continuous("Forecast") +
  facet_wrap(~ lead, ncol = 4) +
  labs(title = "RMSE")


last_plot() %+% skill[measure == "correlation"] + 
  labs(title = "Correlation") +
  geom_cross()
```

## Initial conditions vs forecast

```{r}
skill |> 
  _[measure == "rmse"] |> 
  na.omit() |> 
  ggplot(aes(initial, forecast)) +
  
  geom_point(size = 0.4) +
  geom_smooth(method = "lm", color = "#1c71d8", se = FALSE) +
  # geom_line(data = lines, linewidth = 0.9, color = "#e66100") +
  scale_x_continuous("Initial conditions") +
  scale_y_continuous("Forecast") +
  facet_grid(forecast_season ~ lead) +
  labs(title = "RMSE", 
       subtitle = "As a function of forecast season")


last_plot() %+% skill[measure == "correlation"] + 
  labs(title = "Correlation") +
  geom_cross()
```

## Initial conditions vs forecast


```{r}
skill |> 
  _[measure == "rmse"] |> 
  na.omit() |> 
  ggplot(aes(initial, forecast)) +
  
  geom_point(size = 0.4) +
  geom_smooth(method = "lm", color = "#1c71d8", se = FALSE) +
  # geom_line(data = lines, linewidth = 0.9, color = "#e66100") +
  scale_x_continuous("Initial conditions") +
  scale_y_continuous("Forecast") +
  facet_grid(verification_season ~ lead) +
  labs(title = "RMSE", 
       subtitle = "As a function of verification season")


last_plot() %+% skill[measure == "correlation"] + 
  labs(title = "Correlation") +
  geom_cross()
```

```{r}
skill |> 
  _[measure == "correlation"] |> 
  na.omit() |> 
  ggplot(aes(initial, skill)) +
  geom_point(size = 0.4) +
  geom_smooth(method = "lm", color = "#1c71d8", se = FALSE) +
  # geom_line(data = lines, linewidth = 0.9, color = "#e66100") +
  scale_x_continuous("Initial conditions") +
  scale_y_continuous("Skill") +
  facet_grid(verification_season ~ lead) +
  labs(title = "RMSE skill compared with persistence", 
       subtitle = "As a function of verification season")


last_plot() %+% skill[measure == "correlation"] + 
  labs(title = "Correlation") +
  geom_cross()
```


```{r}
model <- cors[time != time_forecast] |> 
  _[initial, on = c("time_forecast", "member")] |> 
  _[, lead := months_difference(time_forecast, time)] |> 
  _[, let(fisher = atanh(correlation), 
          i.fisher = atanh(i.correlation),
          season = season(time_forecast))] |>  
  lm(fisher ~ season*i.fisher*log(lead) - 1, data = _)


lines <- CJ(i.correlation = seq(-0, 1, length.out = 20),
            lead = seq(1, 8, length.out = 8),
            # month = lubridate::month(1:12, label = TRUE, abbr = TRUE)) |> 
            season = season(c(1, 4, 6, 9))) |>
  _[, let(i.fisher = atanh(i.correlation))] |> 
  
  _[, fisher := predict(model, newdata = .SD)] |> 
  _[, correlation := tanh(fisher)] 
```

## Initial conditions vs forecast

```{r}
cor_plot +
  geom_line(data = lines, linewidth = 1, colour = "#e66100")
```

## Initial conditions vs forecast

```{r}
CJ(i.correlation = .99,
   lead = c(0, seq(1, 8, length.out = 20)),
   season = season(c(1, 4, 6, 9))) |> 
  _[, let(i.fisher = atanh(i.correlation))] |> 
  
  _[, fisher := predict(model, newdata = .SD)] |> 
  _[, correlation := tanh(fisher)] |> 
  _[lead == 0, correlation := i.correlation] |>
  ggplot(aes(lead, correlation)) +
  geom_line(data = \(x) copy(x)[lead != 0][, let(season2 = season, season = NULL)],
            aes(group = season2), color = "gray", alpha =  0.8) +
  geom_line(data = \(x) x[lead != 0]) +
  geom_point(data = \(x) x[lead == 0]) +
  facet_wrap(~season, ncol = 2) +
  scale_x_continuous("Lead time", breaks = 0:12, minor_breaks = NULL) +
  scale_y_continuous("Correlation") +
  labs(caption = "season at forecast time") 

```

## Warning: the linear model is rubbish

```{r}
predicted <- data.table(season = model$model$season,
                        residuals = tanh(model$residuals),
                        predicted = tanh(model$fitted.values))

predicted |> 
  ggplot(aes(predicted, residuals)) + 
  geom_point() +
  geom_smooth() +
  
  predicted |> 
  ggplot(aes(sample = residuals)) + 
  geom_qq() +
  geom_qq_line() &
  
  facet_wrap(~season, ncol = 4) &
  plot_layout(ncol = 1)
```
